Locally-Oriented Programming:\\ A Simple Programming Model for Stencil-Based Computations on Multi-Level Distributed Memory Architectures

### 2015-01-31 papers/LOPe/coarray-comparison.tex
first pass of edits

\item
A programmer is not able to store data to the halo region during execution of a LOPe concurrent
function.  If this were allowed, one thread could overwrite another threads data at undefined times.
The compiler is able to catch this class of error.

@@

\item
A programmer can't make synchronization errors as synchronization is implicit in the
\texttt{CONCURRENT} attribute.  A thread running a concurrent procedure is provided with a copy of
it's local array element plus halo that is consistent with the state of memory \emph{at the time of
  invocation of the procedure}.  Stores to an individual thread's local array element (by that
thread) is never visible to other threads.  LOPE encourages the creation of small functions and lets
the compiler fuse the functions together to improve performance and to provide necessary
synchronization.

>>>>

\item
A programmer is not able to store data to the halo region during
execution of a LOPe concurrent function.  If this were allowed, one
thread could overwrite another threads data at undefined times.  The
compiler is able to catch this class of data race error.

@@

\item
A programmer can't make synchronization errors as synchronization is
implicit in the \texttt{CONCURRENT} attribute\todo{clarify what sync
  errors are prevented.  restricted concurrency model === less room
  for errors.}.  A thread running a concurrent procedure is provided
with a copy of it's local array element plus halo that is consistent
with the state of memory \emph{at the time of invocation of the
  procedure}.  Stores to an individual thread's local array element
(by that thread) is never visible to other threads.  LOPE encourages
the creation of small functions and lets the compiler fuse the
functions together to improve performance and to provide necessary
synchronization.

### 2015-01-31 papers/LOPe/coarray-extensions.tex
first pass of edits

Consider the \texttt{Laplace} concurrent function discussed above.  In this section we demonstrate
how this function can be called in the normal context of a program, one that allows full access to
all of the interior elements of the array, as well as array elements within the logically exterior,
halo-boundary region.  Topics highlighted in this section are: 1. Distributed memory array
allocation; 2. Explicit memory placement; 3. Remote memory transfer; and 4. Remote execution.  This
description is within the context of a Domain Specific Language (DSL), again extending Fortran 2008.
As shorthand, these specific extensions are referred to as CAFe, for Coarray Fortran extensions.

@@

We start with the declaration of an array (to be allocated later),
\begin{verbatim}
  real, allocatable, HALO(1:*:1,1:*:1),          &
        dimension(:,:), codimension[:,:] :: U
\end{verbatim}
with an explicit halo size and with two local dimensions (rank) and two distributed memory
codimensions (corank).  The corank of the array is chosen to be identical to the rank of
the array so that the logical process topology aligns in a way that allows a natural
halo exchange between logically neighboring processes.  For example, if the process
location is \texttt{[pcol,prow]}, then the right-hand halo for the local array \texttt{U}
can be obtained by the assignment \texttt{U(M+1,:) = U(1,:)[pcol+1,prow]} where the size
and cosize of \texttt{U} are given by the allocation statement,
\begin{verbatim}
  allocate( U(0:M+1,0:N+1)[MP,*] )
\end{verbatim}

@@

@@

One obtains a subimage by executing a function call,
\begin{verbatim}
   device = get_subimage(1)
\end{verbatim}
where the integer argument representing one of attached hardware devices.  If the function
fails (e.g., the requested device is unavailable) it returns the image number
\texttt{this\_image()} of the process that is executing the current program.  Returning
the current image allows program execution to proceed correctly even if there are no
attached devices.

@@

Finally, memory for the entire array \texttt{U} can be copied from the subimage device
with the statement,
\begin{verbatim}
  U = U[device]
\end{verbatim}
add memory deallocation is similar to memory allocation:
\begin{verbatim}
   deallocate(U)
   if (device /= this_image()) then
      deallocate(U[device])   [[device]]
   end if
\end{verbatim}

>>>>

Consider the \texttt{Laplace} concurrent function discussed above.  In
this section we demonstrate how this function can be called in the
normal context of a program, one that allows full access to all of the
interior elements of the array, as well as array elements within the
logically exterior, halo-boundary region.  Topics highlighted in this
section are: 1. Distributed memory array allocation; 2. Explicit
memory placement; 3. Remote memory transfer; and 4. Remote execution.
This description is within the context of an extension of the Fortran
2008 language.  As shorthand, these specific extensions are referred
to as CAFe, for Coarray Fortran extensions.

@@

We start with the declaration of an array (to be allocated later),
\begin{verbatim}
  real, allocatable, HALO(1:*:1,1:*:1),          &
        dimension(:,:), codimension[:,:] :: U
\end{verbatim}

with an explicit halo size and with two local dimensions (rank) and
two distributed memory codimensions (corank).  The corank of the array
is chosen to be identical to the rank of the array so that the logical
process topology aligns in a way that allows a natural halo exchange
between logically neighboring processes.\todo{What if corank and rank are not the same?}  For example, if the process
location is \texttt{[pcol,prow]}, then the right-hand halo for the
local array \texttt{U} can be obtained by the assignment
\texttt{U(M+1,:) = U(1,:)[pcol+1,prow]} where the size and cosize of
\texttt{U} are given by the allocation statement,

\begin{verbatim}
  allocate( U(0:M+1,0:N+1)[MP,*] )
\end{verbatim}

@@

\subsection{Subimages}

\todo{subimages are critical and deserve their own subsection}

@@

One obtains a subimage by executing a function call,
\begin{verbatim}
   device = get_subimage(1)
\end{verbatim}

where the integer argument representing one of attached hardware
devices\todo{mention the relation of subimages and hardware earlier in
  defn. of subimage}.  If the function fails (e.g., the requested
device is unavailable) it returns the image number
\texttt{this\_image()} of the process that is executing the current
program.  Returning the current image allows program execution to
proceed correctly even if there are no attached devices.

@@

Finally, memory for the entire array \texttt{U} can be copied from the
subimage device with the statement,

\begin{verbatim}
  U = U[device]
\end{verbatim}

add\todo{Add?  is that typo?} memory deallocation is similar to memory allocation:

\begin{verbatim}
   deallocate(U)
   if (device /= this_image()) then
      deallocate(U[device])   [[device]]
   end if
\end{verbatim}

### 2015-01-31 papers/LOPe/motivation.tex
Completion for coauthors review of sections 2 and 3.
first pass of edits

The concepts in this paper were motivated by problems encountered during
the development of PetaVision, a C++ software framework designed for simulating
large networks of spiking neurons in the visual cortex of primates.  PetaVision
was designed to run on massively parallel hardware architectures and a primitive
version of PetaVision achieved over a Petaflop of sustained single precision
performance on Roadrunner, the fastest computer in the world for a brief
period of time.

@@

TODO: If space allows, give a simple 1D example of connections within a neural network in
Figure 1.  Describe the size of visual cortex in terms of the number of neurons,
the number of synaptic connections, and performance.  Reference SC Gorden Bell
paper from IBM.

>>>>

\todo{This section should be rewritten and remove the petavision stuff}
The concepts in this paper were motivated by problems encountered during
the development of PetaVision, a C++ software framework designed for simulating
large networks of spiking neurons in the visual cortex of primates.  PetaVision
was designed to run on massively parallel hardware architectures and a primitive
version of PetaVision achieved over a Petaflop of sustained single precision
performance on Roadrunner, the fastest computer in the world for a brief
period of time.

@@

TODO: If space allows, give a simple 1D figure of connections within a neural network.
Describe the size of visual cortex in terms of the number of neurons,
the number of synaptic connections, and performance.  Reference SC Gorden Bell
paper from IBM.

### 2015-01-31 papers/LOPe/programming-model.tex
first pass of edits

The LOPe programming model (for Locally-Oriented Programming extensions) restricts the programmer to
a local view of the index space of an array.  Within a LOPe function, only a single array element (called
the local element) is mutable.  In addition, a small halo region surrounding the local element is
visible to the programmer, but this region is immutable.  Restricting the programmer to a local
index space serves to reduce complexity by separating all boundary-value and processor-topology
concerns from the implementation of the algorithm.

@@

LOPe was implemented as an embedded domain specific language (DSL).  Fortran was chosen as the base
language for LOPe, although in principle, the same techniques could be applied to language such as C
or C++ (in conjuction with some form of array extensions such as those proposed by Intel).  Fortran
was chosen because it already provides a rich array-based syntax.  Readers who are unfamiliar with
Fortran syntax may wish to consult Appendix A, for a brief description of Fortran notation.

@@

There are only a few syntax additions required for a LOPe program.  These additions include syntax for
describing halo regions and concurrent procedures.  In all of the code examples that follow, DSL
syntax additions are highlighted by the usage of capitalization for keywords that are either new or
that acquire new usage.

\textbf{Halo regions.}
The principle semantic element of LOPe is the concept of a halo region.  A halo is an ``artificial''
or ``virtual'' region surrounding an array that contains boundary-value information.  Halo (also
called ghost-cell) regions are commonly employed to unify array indexing schemes in the vicinity of
an array boundary, so that an array may be referenced using indices that fall ``outside'' of the
logical domain of the array.  In LOPe, the halo region is given explicit syntax so that the compiler
can exploit this knowledge.  For example, a halo region can be declared with a statement of the form,
\begin{verbatim}
  real, allocatable, dimension(:), HALO(1:*:1) :: A
\end{verbatim}
This statement indicates that \texttt{A} is a rank one array, will be allocated later (when it will
be provided with a specific size), and has a halo region of one element surrounding the array on
either side.  Notation such as \texttt{M:*:N} is to be read as describing a halo region of
\texttt{M} elements to the left, \texttt{N} elements to the right, and with an arbitrary number of
elements in the middle (denoted by the $*$ symbol).  When used to describe a formal parameter of a
function, the halo size may be ``assumed'' (denoted by a ``:'' symbol substituted in place of an
explicit size) as is the following,
\begin{verbatim}
  real, HALO(:,:) :: U
\end{verbatim}
which in this instance indicates that \texttt{U} is a two-dimensional array, with a halo region
whose size is obtained by the compiler from the explicit halo size of the actual array argument
provided at the calling site of the function.  In LOPe, there is no need (in this instance) for a
repetitive \texttt{dimension(:,:)} specification, as it is inferred from the \texttt{HALO}
specification.

\textbf{Concurrent functions.}
The second keyword employed by LOPe is \texttt{concurrent}.  This keyword already exists in the form
of a \texttt{do} \texttt{concurrent} loop construct.  By using this construct, the programmer
asserts that specific iterations of the loop body may be executed by the compiler in \emph{any
  order;} even \emph{concurrently.}
%%LOPe allows the \texttt{concurrent} attribute to be attached to a function declaration so that the function can be called from within a concurrent loop without fear of race conditions.
For example, the following,
\begin{verbatim}
  pure CONCURRENT subroutine Laplace(U)
     real, HALO(:,:) :: U
     U(0,0) =                 U(0,+1)               &
              +  U(-1,0)  - 3*U(0, 0)  +  U(+1,0)   &
                          +   U(0,-1)
  end subroutine Laplace
\end{verbatim}
defines a function that can be used as part of an iterative solution of Laplace's equation in two
dimensions.  LOPe allows a function with the \texttt{pure} and \texttt{concurrent} attributes to be
called from within a \texttt{do} \texttt{concurrent} loop body.  One should imagine that this
function is called \emph{for each} \texttt{(i,j)} index of the interior of the array \texttt{U}.
(An example of this usage will be given later in the text.)  Note that this usage introduces a race
condition as new values of elements of \texttt{U} are created on the left-hand side of the
assignment statement that may use \emph{new or old} values of \texttt{U} on the right-hand side.
LOPe requires the compiler to guarantee that race conditions won't occur by using, e.g.,
double-buffering techniques as needed.

\textbf{LOPe index notation.}
In the \texttt{Laplace} example shown above, the \texttt{U(0,0)} array element is the \emph{local}
array element and only the local element may be modified.  This zero-based indexing for the
local-array element is different than conventional Fortran, where by default, array indices start at
1.  For LOPe, this default convention is modified; indices for arrays declared with the
\texttt{HALO} attribute (within the context of a \texttt{CONCURRENT} procedure) range (in one
dimension) from \texttt{-M} to \texttt{+N} where the halo attribute has been declared as
\texttt{M:*:N} by the originating type declaration.  The other array elements (in the halo region of
the example) are \texttt{U(-1,0)} and \texttt{U(+1,0)} (left and right of local, respectively) and
\texttt{U(0,-1)} and \texttt{U(0,+1)} (below and above of local, respectively).  The geometric
positioning of the array elements can be seen by examining the arrangement of the expressions on the
right-hand side of the \texttt{Laplace} example.

>>>>

The LOPe programming model (for Locally-Oriented Programming
extensions) restricts the programmer to a local view of the index
space of an array.  Within a LOPe function, only a single array
element (called the local element) is mutable.  In addition, a small
halo region surrounding the local element is visible to the
programmer, but this region is immutable.  Restricting the programmer
to a local index space serves to reduce complexity by separating all
data- and task-decomposition concerns from the implementation of the
element-level array calculations.

@@

LOPe is a domain specific language (DSL) implemented as a small extension
to standard Fortran 2008\todo{What version of Fortran does it extend?}.
Fortran was chosen as the base language for LOPe, although in
principle, the same techniques could be applied to language such as C
or C++ (in conjuction with some form of array extensions such as those
proposed by Intel).  Fortran was chosen because it already provides a
rich array-based syntax.  Readers who are unfamiliar with Fortran
syntax may wish to consult Appendix A, for a brief description of
Fortran notation.

\subsection{Related work}

LOPe builds upon prior work studying how to map Fortran to accelerator
programming models like OpenCL.  In the ForOpenCL project~\cite{foropencl}
we exploited Fortran pure and elemental functions to express the core
data parallel kernels of an OpenCL-based program.  In practice, array
calculations for a given index $i,j$ will require read-only access to
a local neighborhood of size $[M,N]$ around $i,j$.  The LOPe work introduces
a mechanism for representing these neighborhoods as array declaration
type annotations.

This work is based on concepts explored in the ZPL programming
language~\cite{zpl1,zpl2} in which the programmer can define regions and 
operators that are applied over the index sets corresponding to the regions.
For example, the following ZPL code implements the same stencil as the
LOPe code in Section~\ref{sec:where-is-jacobi-example}:

\begin{verbatim}
ZPL jacobi example here.
\end{verbatim}

This approach as demonstrated by ZPL is quite powerful for compilation
purposes since it provides a clean decoupling of the operators applied
over an array from the decomposition of that array over a potentially
complex distributed memory hierarchy.

@@

There are only a few syntax additions required for a LOPe program.
These additions include syntax for describing halo regions and
concurrent procedures.  In all of the code examples that follow,
syntax additions are highlighted by the usage of capitalization for
keywords that are either new or that acquire new usage.

\textbf{Halo regions.}

The principle semantic element of LOPe is the concept of a halo
region.  A halo is an ``artificial'' or ``virtual'' region surrounding
an array that contains boundary-value information.  Halo (also called
ghost-cell) regions are commonly employed to unify array indexing
schemes in the vicinity of an array boundary, so that an array may be
referenced using indices that fall ``outside'' of the logical domain
of the array.  In LOPe, the halo region is given explicit syntax so
that the compiler can exploit this information for purposes of memory
allocation, data replication and thread synchronization.  For example,
a halo region can be declared with a statement of the form,

\begin{verbatim}
  real, allocatable, dimension(:), HALO(1:*:1) :: A
\end{verbatim}

This statement indicates that \texttt{A} is a rank one array, will be
allocated later (when it will be provided with a specific size), and
has a halo region of one element surrounding the array on either side.
Notation such as \texttt{M:*:N} is to be read as describing a halo
region of \texttt{M} elements to the left, \texttt{N} elements to the
right, and with an arbitrary number of elements in the middle (denoted
by the $*$ symbol).  When used to describe a formal parameter of a
function, the halo size may be ``assumed'' (denoted by a ``:'' symbol
substituted in place of an explicit size) as is the following,

\begin{verbatim}
  real, HALO(:,:) :: U
\end{verbatim}

which in this instance indicates that \texttt{U} is a two-dimensional
array, with a halo region whose size is obtained by the compiler from
the explicit halo size of the actual array argument provided at the
calling site of the function.  In LOPe, there is no need (in this
instance) for a repetitive \texttt{dimension(:,:)} specification, as
it is inferred from the \texttt{HALO} specification.\todo{This paragraph is a bit weird to read.}

\textbf{Concurrent functions.}
The second keyword employed by LOPe is \texttt{concurrent}.  This keyword already exists in the form
of a \texttt{do} \texttt{concurrent} loop construct.  By using this construct, the programmer
asserts that specific iterations of the loop body may be executed by the compiler in \emph{any
  order;} even \emph{concurrently.}
%%LOPe allows the \texttt{concurrent} attribute to be attached to a function declaration so that the function can be called from within a concurrent loop without fear of race conditions.
For example, the following,
\begin{verbatim}
  pure CONCURRENT subroutine Laplace(U)
     real, HALO(:,:) :: U
     U(0,0) =                 U(0,+1)               &
              +  U(-1,0)  - 3*U(0, 0)  +  U(+1,0)   &
                          +   U(0,-1)
  end subroutine Laplace
\end{verbatim}

defines a function that can be used as part of an iterative solution
of Laplace's equation in two dimensions.  LOPe allows a function with
the \texttt{pure} and \texttt{concurrent} attributes to be called from
within a \texttt{do} \texttt{concurrent} loop body.\todo{Worth pointing out that pure implies no side effects, concurrent implies no dependencies between iterations?}  One should
imagine that this function is called \emph{for each} \texttt{(i,j)}
index of the interior of the array \texttt{U}.  (An example of this
usage will be given later in the text.)  Note that this usage
introduces a race condition as new values of elements of \texttt{U}
are created on the left-hand side of the assignment statement that may
use \emph{new or old} values of \texttt{U} on the right-hand side.
LOPe requires the compiler to guarantee that race conditions won't
occur by using, e.g., double-buffering techniques as needed.

\textbf{LOPe index notation.}

In the \texttt{Laplace} example shown above, the \texttt{U(0,0)} array
element is the \emph{local} array element and only the local element
may be modified.  This zero-based indexing for the local-array element
is different than conventional Fortran, where by default, array
indices start at 1.  The use of zero-based indexing gives a clean
symmetry for indices on either side of the central element at zero.
For LOPe, this default convention is modified; indices for arrays
declared with the \texttt{HALO} attribute (within the context of a
\texttt{CONCURRENT} procedure) range (in one dimension) from
\texttt{-M} to \texttt{+N} where the halo attribute has been declared
as \texttt{M:*:N} by the originating type declaration.  The other
array elements (in the halo region of the example) are
\texttt{U(-1,0)} and \texttt{U(+1,0)} (left and right of local,
respectively) and \texttt{U(0,-1)} and \texttt{U(0,+1)} (below and
above of local, respectively).  The geometric positioning of the array
elements can be seen by examining the arrangement of the expressions
on the right-hand side of the \texttt{Laplace} example.
